### 2018-04-28 doc/paper.tex
Adding write-up to section about existing algorithm.
Though I highly seasoned the original compensated de Casteljau
paper with my own notation.

\begin{theorem}[\cite{Ogita2005}]
For \(a, b \in \mathbf{F}\) and \(x, y \in \mathbf{F}\), \texttt{TwoSum} and
\texttt{TwoProd} satisfy
\begin{itemize}
  \item \(\left[x, y\right] = \mathtt{TwoSum}(a, b)\), \(x = \fl{a + b}\),
      \(x + y = a + b\), \(y \leq \mach \left|x\right|\),
      \(y \leq \mach \left|a + b\right|\)
  \item \(\left[x, y\right] = \mathtt{TwoProd}(a, b)\),
      \(x = \fl{a \times b}\), \(x + y = a \times b\),
      \(y \leq \mach \left|x\right|\),
      \(y \leq \mach \left|a \times b\right|\).
\end{itemize}
\end{theorem}

@@

  \begin{algorithmic}
    \Function{\(\mathtt{res} = \mathtt{DeCasteljau}\)}{$\mathbf{b}, s$}
      \State \(n = \texttt{length}(b) - 1\)
      \State \(r = 1 \ominus s\)
      \\
      \For{\(j = 0, \ldots, n\)}
        \State \(b_j^{(n)} = b_j\)
      \EndFor
      \\
      \For{\(k = n - 1, \ldots, 0\)}
        \For{\(j = 0, \ldots, k\)}
          \State \(b_j^{(k)} = \left(r \otimes b_j^{(k + 1)}\right) \oplus
              \left(s \otimes b_{j + 1}^{(k + 1)}\right)\)
        \EndFor
      \EndFor
      \\
      \State \(\mathtt{res} = b_0^{(0)}\)
    \EndFunction
  \end{algorithmic}
\end{algorithm}

@@

@@

  \begin{algorithmic}
    \Function{\(\mathtt{res} = \mathtt{CompDeCasteljau}\)}{$\mathbf{b}, s$}
      \State \(n = \texttt{length}(\mathbf{b}) - 1\)
      \State \(\left[r, \rho\right] = \mathtt{TwoSum}(1, -s)\)
      \\
      \For{\(j = 0, \ldots, n\)}
        \State \(b_j^{(n)} = b_j\)
        \State \(\partial b_j^{(n)} = 0\)
      \EndFor
      \\
      \For{\(k = n - 1, \ldots, 0\)}
        \For{\(j = 0, \ldots, k\)}
          \State \(\left[P_1, \pi_1\right] = \mathtt{TwoProd}\left(
              r, b_j^{(k + 1)}\right)\)
          \State \(\left[P_2, \pi_2\right] = \mathtt{TwoProd}\left(
              s, b_{j + 1}^{(k + 1)}\right)\)
          \State \(\left[b_j^{(k)}, \sigma_3\right] = \mathtt{TwoSum}(
              P_1, P_2)\)
          \State \(w = \pi_1 \oplus \pi_2 \oplus \sigma_3 \oplus
              \left(\rho \otimes b_j^{(k + 1)}\right)\)
          \State \(\partial b_j^{(k)} = w \oplus \left(s \otimes
              \partial b_{j + 1}^{(k + 1)}
              \right) \oplus \left(r \otimes \partial b_j^{(k + 1)}\right)\)
        \EndFor
      \EndFor
      \\
      \State \(\mathtt{res} = b_0^{(0)} \oplus \partial b_0^{(0)}\)
    \EndFunction
  \end{algorithmic}
\end{algorithm}

@@

Consider de Casteljau's method to evaluate a degree \(n\)
polynomial in Bernstein-B\'{e}zier form with control points \(b_j\):
\begin{align*}
    b_j^{(n)} &= b_j \\
    b_j^{(k)} &= (1 - s) b_j^{(k + 1)} + s b_{j + 1}^{(k + 1)} \\
    b(s) &= b_0^{(0)}.
\end{align*}

>>>>

\begin{theorem}[\cite{Ogita2005}]
For \(a, b \in \mathbf{F}\) and \(P, \pi, S, \sigma \in \mathbf{F}\),
\texttt{TwoSum} and \texttt{TwoProd} satisfy
\begin{itemize}
  \item \(\left[S, \sigma\right] = \mathtt{TwoSum}(a, b)\), \(x = \fl{a + b}\),
      \(S + \sigma = a + b\), \(\sigma \leq \mach \left|S\right|\),
      \(\sigma \leq \mach \left|a + b\right|\)
  \item \(\left[P, \pi\right] = \mathtt{TwoProd}(a, b)\),
      \(P = \fl{a \times b}\), \(P + \pi = a \times b\),
      \(\pi \leq \mach \left|P\right|\),
      \(\pi \leq \mach \left|a \times b\right|\).
\end{itemize}
The letters \(\sigma\) and \(\pi\) are used to indicate that the
errors came from sum and product, respectively.
\end{theorem}

@@

  \begin{algorithmic}
    \Function{\(\mathtt{res} = \mathtt{DeCasteljau}\)}{$b, s$}
      \State \(n = \texttt{length}(b) - 1\)
      \State \(\widehat{r} = 1 \ominus s\)
      \\
      \For{\(j = 0, \ldots, n\)}
        \State \(\widehat{b}_j^{(n)} = b_j\)
      \EndFor
      \\
      \For{\(k = n - 1, \ldots, 0\)}
        \For{\(j = 0, \ldots, k\)}
          \State \(\widehat{b}_j^{(k)} = \left(
              r \otimes \widehat{b}_j^{(k + 1)}\right) \oplus
              \left(s \otimes \widehat{b}_{j + 1}^{(k + 1)}\right)\)
        \EndFor
      \EndFor
      \\
      \State \(\mathtt{res} = \widehat{b}_0^{(0)}\)
    \EndFunction
  \end{algorithmic}
\end{algorithm}

@@

In this section we review the compensated de Casteljau algorithm
from \cite{Jiang2010}. In order to track the local errors at
each update step, we use four EFTs:
\begin{align}
\left[\widehat{r}, \rho\right] &= \mathtt{TwoSum}(1, -s) \\
\left[P_1, \pi_1\right] &= \mathtt{TwoProd}\left(\widehat{r}, \widehat{b}_j^{(k + 1)}\right) \\
\left[P_2, \pi_2\right] &= \mathtt{TwoProd}\left(s, \widehat{b}_{j + 1}^{(k + 1)}\right) \\
\left[\widehat{b}_j^{(k)}, \sigma_3\right] &= \mathtt{TwoSum}(P_1, P_2)
\end{align}
With these, we can exactly describe the local error between the exact
update and computed update:
\begin{gather}
\ell_j^{(k)} = \pi_1 + \pi_2 + \sigma_3 + \rho\widehat{b}_j^{(k + 1)} \\
(1 - s) \cdot \widehat{b}_j^{(k + 1)} +
  s \cdot \widehat{b}_{j + 1}^{(k + 1)} =
\widehat{b}_j^{(k)} + \ell_j^{(k)}.
\end{gather}

\begin{figure}[H]
\centering
\begin{tikzpicture}[node distance = 0.2\textwidth]
  \node (b0) {\(\widehat{b}^{(n)}\)}
    child {node (b1) {\(\widehat{b}^{(n - 1)}\)}
      child {node (b2) {\(\widehat{b}^{(n - 2)}\)}
        child {node (bdots) {\(\vdots\)}
          child {node (bn) {\(\widehat{b}^{(0)}\)}}
          child {node (ln) {\(\ell^{(0)}\)}}
        }
        child {node (ldots) {\(\vdots\)}}
      }
      child {node (l2) {\(\ell^{(n - 2)}\)}}
    }
    child {node (l1) {\(\ell^{(n - 1)}\)}};
  %%
  \path (b1) -- (l1) node [midway] {\(+\)};
  \path (b2) -- (l2) node [midway] {\(+\)};
  \path (bn) -- (ln) node [midway] {\(+\)};
  %% eb == exact b
  \node [right of=b0] (eb0) {\(b^{(n)}\)}
    child {node (eb1) {\(b^{(n - 1)}\)}
      child {node (eb2) {\(b^{(n - 2)}\)}
        child {node (ebdots) {\(\vdots\)}
          child {node (ebn) {\(b^{(0)}\)}}
        }
      }
    };
\end{tikzpicture}
\caption{Local round-off errors}
\end{figure}

\noindent By defining the global errors at each step
\begin{equation}
  \partial b_j^{(k)} = b_j^{(k)} - \widehat{b}_j^{(k)}
\end{equation}
we can see that the local errors accumulate in
\(\partial b^{(k)}\):
\begin{equation}
  \partial b_j^{(k)} = (1 - s) \cdot \partial b_j^{(k + 1)} + s \cdot
  \partial b_{j + 1}^{(k + 1)} + \ell_j^{(k)}.
\end{equation}
Hence, when computed in exact arithmetic, we have
\begin{equation}
  p(s) = \widehat{b}_0^{(0)} + \partial b_0^{(0)}.
\end{equation}

The idea behind the compensated de Casteljau algorithm
is to compute both the local error and the updates of the global
error with floating point operations:

  \begin{algorithmic}
    \Function{\(\mathtt{res} = \mathtt{CompDeCasteljau}\)}{$b, s$}
      \State \(n = \texttt{length}(b) - 1\)
      \State \(\left[\widehat{r}, \rho\right] = \mathtt{TwoSum}(1, -s)\)
      \\
      \For{\(j = 0, \ldots, n\)}
        \State \(\widehat{b}_j^{(n)} = b_j\)
        \State \(\widehat{\partial b}_j^{(n)} = 0\)
      \EndFor
      \\
      \For{\(k = n - 1, \ldots, 0\)}
        \For{\(j = 0, \ldots, k\)}
          \State \(\left[P_1, \pi_1\right] = \mathtt{TwoProd}\left(
              \widehat{r}, \widehat{b}_j^{(k + 1)}\right)\)
          \State \(\left[P_2, \pi_2\right] = \mathtt{TwoProd}\left(
              s, \widehat{b}_{j + 1}^{(k + 1)}\right)\)
          \State \(\left[\widehat{b}_j^{(k)}, \sigma_3\right] =
              \mathtt{TwoSum}(P_1, P_2)\)
          \State \(\widehat{\ell}_j^{(k)} = \pi_1 \oplus \pi_2 \oplus \sigma_3
              \oplus \left(\rho \otimes \widehat{b}_j^{(k + 1)}\right)\)
          \State \(\widehat{\partial b}_j^{(k)} =
              \left(\widehat{r} \otimes
              \widehat{\partial b}_j^{(k + 1)}\right) \oplus
              \left(s \otimes \widehat{\partial b}_{j + 1}^{(k + 1)}
              \right) \oplus
              \widehat{\ell}_j^{(k)}\)
        \EndFor
      \EndFor
      \\
      \State \(\mathtt{res} = \widehat{b}_0^{(0)} \oplus
          \widehat{\partial b}_0^{(0)}\)
    \EndFunction
  \end{algorithmic}
\end{algorithm}

\noindent  When comparing this computed error to the exact error, the
difference depends only on \(s\) and the Bernstein
coefficients of \(p\):

\begin{theorem}[\cite{Jiang2010}]
  If no underflow occurs and \(n \geq 2\), then
  \begin{equation}
    \left|\partial b_0^{(0)} - \widehat{\partial b}_0^{(0)}\right| \leq
    2 \gamma_{3n + 1} \gamma_{3(n - 1)}
    \sum_{j = 0}^n \left|b_j\right| B_{j, n}(s).
  \end{equation}
\end{theorem}

\noindent Using the bound on the error in \(\partial b\), the algorithm can
be shown to be as accurate as if the computations were done in twice
the working precision:

\begin{theorem}[\cite{Jiang2010}]
  If no underflow occurs, \(n \geq 2\) and \(s \in \left[0, 1\right]\)
  \begin{equation}
    \frac{\left|p(s) - \mathtt{CompDeCasteljau}(p, s)\right|}{\left|p(s)\right|} \leq \mach +
    2 \gamma_{3n}^2 \cond{p, s}.
  \end{equation}
\end{theorem}

@@

@@

Consider de Casteljau's method to evaluate a degree \(n\)
polynomial in Bernstein-B\'{e}zier form with control points \(b_j\):
\begin{align}
    b_j^{(n)} &= b_j \\
    b_j^{(k)} &= (1 - s) b_j^{(k + 1)} + s b_{j + 1}^{(k + 1)} \\
    b(s) &= b_0^{(0)}.
\end{align}
